Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-05-13 figures.Rmd
code for Figures_draft2

load("model1.Rdata")
load("model2.Rdata")

@@

```{r fig1}

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Deoxy Hb`, `Deoxy Hb α`, `Oxy Hb`, `Oxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
  dplyr::filter(`Deoxy Hb α` < 3)

@@

**Figure 1.** Differences in variables measured by Oxiplex and associated α values among patient groups. THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033). Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07). Note: removed one outlier from Deoxy Hb α (3.52)

### Figure 3: Logistic regression on CM vs UM

@@

```{r fig3a}
# select variables of interest
df_hbox_log_voi <- df_hbox_impute %>% 
  filter(Status != "HV") %>% 
  dplyr::select(2, 11:19)

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_voi %>% 
  dplyr::select(-Status) %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  as.data.frame() 
rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id

# compute PCA
my_pca2 <- prcomp(df_hbox_voi_df, scale = TRUE)

# define groups
groups2 <- as.factor(df_hbox_log_voi$Status)

# Eigenvalues
pca_eig_val2 <- get_eigenvalue(my_pca2)

# two data frames for plotting
df_vars2 <- my_pca2$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca2$rotation))
df_pnts2 <- my_pca2$x %>% as_tibble() %>% mutate(group = groups2)

@@

  # plot patients
  ggplot() +
  geom_point(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars2, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars2, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars2, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

@@

  # design
  xlab(paste0("PC1: ", round(pca_eig_val2$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val2$variance.percent[2], 2), "% of variance")) +

@@

```{r fig3b_prep, eval = FALSE}

@@

# select only significant variables
df_hbox_model2 <- df_hbox_voi_01 %>% 
  dplyr::select(c(Status, THC_alpha, THC_sd)) %>% 
  dplyr::rename("THCα" = "THC_alpha", "THC sd" = "THC_sd") %>% 
  mutate(`1/10 * THCα` = THCα/10) %>% 
  dplyr::select(-THCα)

# resample
set.seed(789)
hbox_boot2 <- bootstraps(df_hbox_model2, times = 1000)

# recipe
hbox_rec2 <- recipe(Status ~ ., data = df_hbox_model2) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# workflow
hbox_wf2 <- workflow() %>% 
  add_recipe(hbox_rec2)

# logistic regression model spec
glm_spec2 <- logistic_reg() %>% 
  set_engine("glm")

glm_res2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>%
  fit_resamples(
    resamples = hbox_boot2,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
glm_best2 <- glm_res2 %>% select_best("roc_auc")

# finalize model
hbox_final2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>% 
  finalize_workflow(glm_best2) %>%
  fit(df_hbox_model2) %>%
  pull_workflow_fit()

# save objects so model doesn't rerun 
save(glm_res2, hbox_final2, file = "model2.Rdata")

@@

```{r fig3b_plot}

final_res_conf_mat2 <- glm_res2 %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM"))

final_res_conf_mat2 %>% 
  group_by(Truth) %>% 
  summarise(total_true = sum(Freq)) %>% 
  right_join(final_res_conf_mat2) %>% 
  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%

@@

**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα and THC sd as predictors. **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred.

@@

collect_metrics(glm_res2) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

hbox_final2 %>% 
  tidy() %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Supplementary figure 1: Variable correlations 
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

@@

df_hbox_cor <- df_hbox_impute %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  dplyr::select(Status, `Deoxy Hb`, `Oxy Hb`, `O2 sat`, THC, `THC sd`, `Deoxy Hb α`, `Oxy Hb α`, `O2 sat α`, THCα)

m_cor_hv <- df_hbox_cor %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_cor %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_cor %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")) %>%

@@

    xlab("") +
    ylab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
          axis.text = element_text(size = 35),
          legend.text = element_text(size = 40),
          legend.title = element_text(size = 50),
          legend.key.size = unit(2, 'cm'),
          plot.title = element_text(size = 30)
    )

@@

### Supplementary figure 2: Logistic regression curves of NIRS variables
```{r fig_s2}

@@

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x") +

@@

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher disciminatory power

>>>>

load("glm_model.Rdata")

@@

```{r fig1, fig.height = 12}

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
  dplyr::filter(`Deoxy Hb α` < 3)

@@

**Figure 1.** Differences in variables measured by Oxiplex and associated α values among patient groups. THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033). Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07). Note: removed one outlier from Deoxy Hb α (3.52).

### Figure 2: Logistic regression on CM vs UM

@@

```{r fig2a}
# remove healthy volunteers, Oxy Hb, and Deoxy Hb from analysis
df_hbox_log_voi <- df_hbox_voi %>% 
  dplyr::filter(Status != "HV") %>% 
  dplyr::select(-(2:5))

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_voi %>% 
  dplyr::select(-Status) %>% 
  as.data.frame() 
rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id

# compute PCA
my_pca <- prcomp(df_hbox_voi_df, scale = TRUE)

# define groups
groups <- as.factor(df_hbox_log_voi$Status)

# Eigenvalues
pca_eig_val <- get_eigenvalue(my_pca)

# two data frames for plotting
df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)

@@

  # plot patients
  ggplot() +
  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), size = 3) +
  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

@@

  # design
  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +

@@

```{r fig2b_prep, eval = FALSE}

@@

# select only significant variables
df_hbox_model <- df_hbox_voi_01 %>% 
  dplyr::select(c(Status, THCα)) %>% 
  mutate(`1/10 * THCα` = THCα/10) %>% 
  dplyr::select(-THCα)

# resample
set.seed(123)
hbox_boot <- bootstraps(df_hbox_model, times = 1000)

# recipe
hbox_rec <- recipe(Status ~ ., data = df_hbox_model) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# create workflow
hbox_wf <- workflow() %>% add_recipe(hbox_rec)

# logistic regression model specifications
glm_spec <- logistic_reg() %>% set_engine("glm")

# create model on 1000 different resamples
glm_res <- hbox_wf %>% 
  add_model(glm_spec) %>%
  fit_resamples(
    resamples = hbox_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
glm_best <- glm_res %>% select_best("roc_auc")

# finalize model
hbox_final <- hbox_wf %>% 
  add_model(glm_spec) %>% 
  finalize_workflow(glm_best) %>%
  fit(df_hbox_model) %>%
  pull_workflow_fit()

# save objects so model doesn't rerun 
save(glm_res, hbox_final, file = "glm_model.Rdata")

@@

```{r fig2b_plot}

final_res_conf_mat <- glm_res %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM"))

final_res_conf_mat %>% 
  group_by(Truth) %>% 
  summarise(total_true = sum(Freq)) %>% 
  right_join(final_res_conf_mat) %>% 
  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%

@@

**Figure 2** Logistic regression model discriminating CM vs UM patients using only THCα as the predictor. **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.

@@

collect_metrics(glm_res) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

hbox_final %>% 
  tidy() %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Supplementary figure 1: Spearman correlations of NIRS variables 
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

@@

m_cor_hv <- df_hbox_voi %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_voi %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_voi %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC", "Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%

@@

    xlab("") +
    ylab("") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
          axis.text = element_text(size = 35),
          legend.text = element_text(size = 40),
          legend.title = element_text(size = 50),
          legend.key.size = unit(2, 'cm'),
          plot.title = element_text(size = 30)
    )

@@

### Supplementary figure 2: Logistic regression curves of NIRS variables
```{r fig_s2, fig.height = 12}

@@

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +

@@

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher discriminatory power
